---
title: "Career Advancement"
author: "Blondie"
format: pdf
---

```{r}
#Load All libraries
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(stringr)
library(lubridate)
library(tidyr)
library(car)
library(janitor)
library(caret)
library(glmnet)
library(dplyr)
library(tigris)
library(ggplot2)
library(sf)
library(forcats)
library(kableExtra)
```

## Load all datasets and clean column names to snake_case with janitor
```{r}
# Loading and Processing 
intake <- read_excel("Intake Survey _AU2.xlsx") |> clean_names()
training_edu <- read_excel("Training&Education_AU.xlsx") |> clean_names()
employment <- read_excel("Employment Data_AU2.xlsx") |> clean_names()
july_data <- read_excel("July 25 Monthly Data_AU.xlsx") |> clean_names()
```

# INTAKE DATASET
```{r}
# Drop columns with > 60% missing in Intake data
threshold <- 0.6 * nrow(intake)
intake <- intake |> 
  select(where(~ sum(is.na(.)) < threshold))

# Select and change some variables into factors 
intake <- intake |> 
  select(response_id, education_goals, training_completed, desired_career_interests) |>
  mutate(across(where(is.character), as.factor)) |>
  mutate(across(where(is.factor), ~ fct_explicit_na(., na_level = "Unknown")))
```

## Program Intake Education Goals
```{r}
# Define keywords for each education level
education_patterns <- list(
  "High School/GED" = c("high school", "ged"),
  "Associate Degree" = c("associate"),
  "Bachelor Degree" = c("bachelor"),
  "Masters or Higher" = c("master", "phd", "doctorate"),
  "Vocational Certifications" = c("vocational"),
  "Occupational License" = c("occupational license"),
  "None (Already Have What I Need)" = c("none \\(i already have", "i'm not sure\\r\\nnone"),
  "Unknown" = c("unknown", "other", "i'm not sure", "im not sure")
)

# Collapse patterns into regex for easier matching
education_patterns <- lapply(education_patterns, function(x) paste(x, collapse = "|"))

# Apply categorization
intake <- intake |> 
  mutate(
    education_goals_clean = tolower(education_goals),
    education_goals = purrr::map_chr(education_goals_clean, ~ {
      matched <- names(education_patterns)[sapply(education_patterns, function(pat) str_detect(.x, pat))]
      if (length(matched) == 0) "Other" else matched[1]
    })
  ) |> 
  select(-education_goals_clean)
```

```{r,}
#Distribution of education Goal Category
intake |> 
  count(education_goals, sort = TRUE) |> 
  ggplot(aes(x = fct_reorder(education_goals, n), y = n)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(position='right') +
  coord_flip() +
  labs(title = "Distribution of Education Goals at Intake",
       x = "",
       y = "")
```

# Program Intake Training Completed 
```{r}
#Training Completed
training_patterns <- list(
  "None" = c("none", "no training", "n/a", "na", "have not completed any yet", "no"),
  
  #Education levels 
  "High School Diploma / GED" = c("high school", "ged", "diploma"),
  "Associate Degree" = c("associate"),
  "Bachelor's Degree" = c("bachelor", "ba", "bs"),
  "Masters or Higher" = c("master", "m\\.a", "msc", "phd", "doctorate", "mba"),
  
  # Healthcare & Social Services
  "Healthcare" = c("medical", "dental", "cpr", "aid", "nurse","cna","health","clinical",
                   "phlebotomy","emt","patient care","bls","hipaa","serv safe","crp",
                   "safe certification blood pathogen", "pharmacy technician", 
                   "pharmacy", "nursing", "care giver"),
  #Legal 
  "Legal" = c("attorney","paralegal","judge","legal","court","compliance","notary","mediation"),
  
   # Business, Entrepreneurship & Office
  "Business & Entrepreneurship" = c("entrepreneur","business","management","project","retail","finance"),
  
   # IT & Digital
  "Information Technology" = c("it","computer","cyber","network","compTIA",
                               "software","programming","digital","comptia a","help desk",
                               "mos certification specialist in powerpoint, excel and word"),
  
    # Hospitality & Food Service
  "Hospitality & Food Service" =c("hospitality","food","chef","cook",
                                  "restaurant","bartend","catering","hotel","serve safe"),
  
  # Childcare & Education
  "Childcare & Education" = c("child","childcare","teacher","ece","daycare",
                              "early childhood","paraprofessional"),
  
    # Customer Service
  "Customer Service & Admin" = c("customer service","clerk","admin","office",
                                 "front desk","receptionist","sales","cashier",
                                 "conceirge","concierge"),
  
   # Trades, Construction & Logistics
  "Construction & Technical Trades" = c("construction","carpentry","electric","maintenance",
                                        "plumbing","truck","forklift"),
   # Cosmetology & Personal Care
  "Cosmetology & Personal Care" = c("cosmetology","esthetic","hair","barber","beauty","spa",
                                    "makeup","nail","waxing","facial","eyebrow"),
  
  #Social work and 
  "Social Work & Human Services" = c("social work","counseling","mental health","case management","community support","trauma","advocacy","homeless", "youth"   ),
  
   # Security & Law Enforcement
  "Security & Law Enforcement" = c("security","police","armed","unarmed","law enforcement","military","firearm","spo"),
  
  #transport and logistics 
  "Transportation & Logistics" = c("cdl","truck","driver","delivery","transport",
                                   "warehouse","logistics","route","pallet jack"),
  
  "Office & Professional Development" = c("training","certificate","certified","leadership",
                                          "management","program","development","dc career",
                                          "job corps","fema", "apprentice", "dhl"),
  # Creative & Specialized Fields
  "Creative & Specialized Fields" = c("music","yoga","art","design","creative","photography","writing","media"),
  
  #In Education or Training
  "In Education or Training" = c("student","enrolled","college","university","learning","course","class","in the process","currently studying"),
  
  #General Certified Skills / Licensing
  "General Certified Skills" = c("certificate","certified","license","licensure","accreditation","credential")
)

# Convert keywords to regex for str_detect
training_patterns <- lapply(training_patterns, function(x) paste(x, collapse = "|"))

# Categorize training_completed
intake <- intake |> 
  mutate(
    training_category = map_chr(training_completed, function(tc) {
      tc_lower <- tolower(tc)
      matched <- names(training_patterns)[sapply(training_patterns, function(pat) str_detect(tc_lower, pat))]
      if(length(matched) == 0) "Uncategorized" else matched[1]
    })
  )

# Remove duplicates
intake_unique <- intake |> 
  distinct(training_completed, training_category)
```

```{r}
#Visualization of the Training Completed variable
intake |> 
  count(training_category, sort = TRUE) |> 
  ggplot(aes(x = fct_reorder(training_category, n), y = n)) +
  geom_col(fill = "darkgreen") +
  scale_y_continuous(position='right') +
  coord_flip() +
  labs(title = "Distribution of Training Completed by Category",
       x = "",
       y = "") +
  theme(panel.grid.major.y = element_blank())
```

```{r}
# Get all entries that were not uncategorized. 
uncategorized_levels <- intake |> 
  filter(training_category == "Uncategorized") |> 
  pull(training_completed) |>  
  unique()                       

# View them
uncategorized_levels
```

## Desired Career Goals. 
```{r}
#Desired Career Goals at Intake
# Categorize career interests into standardized clusters using keyword
intake <- intake |> 
  mutate(
    desired_careers = case_when(
      str_detect(str_to_lower(desired_career_interests), 
                 "unknown|not sure|i'm not sure|im not sure") ~ "Unknown",
      str_detect(str_to_lower(desired_career_interests), 
                 "business|administrative|management|admin") ~
        "Administrative",
      str_detect(str_to_lower(desired_career_interests), 
                 "construction|trades|carpentry|electric|plumbing|welding|
                 cdl|forklift") ~ "Specialized Trades",
      str_detect(str_to_lower(desired_career_interests), "education|teacher|
                 teaching|early childhood|school") ~ "Education",
      str_detect(str_to_lower(desired_career_interests), "entrepreneur|
                 start your own|small business|
                 self employment") ~ "Start Own Business",
      str_detect(str_to_lower(desired_career_interests), 
                 "government|public sector|civil service") ~ "Government",
      str_detect(str_to_lower(desired_career_interests), "healthcare|nurse|
                 medical|doctor|pharmacy|health") ~ "Healthcare",
      str_detect(str_to_lower(desired_career_interests), "hospitality|hotel|
                 restaurant|food service|bartender|waiter|serve safe food handlers") ~ "Hospitality",
      str_detect(str_to_lower(desired_career_interests), "human services|
                 social work|counseling|community") ~ 
        "Human Services/ Social Work",
      str_detect(str_to_lower(desired_career_interests), 
                 "information technology|it|computer|cybersecurity|software|microsoft|
                 tech") ~ "Information Technology",
      str_detect(str_to_lower(desired_career_interests), "infrastructure|
                 utility|solar|energy") ~ "Infrastructure and energy",
      str_detect(str_to_lower(desired_career_interests), "science|technology|
                 engineering|math|stem") ~ "Science and Technology",
      str_detect(str_to_lower(desired_career_interests), "security|
                 law enforcement|police|military|firefighter") ~ 
        "Security and Law enforcement",
      str_detect(str_to_lower(desired_career_interests), "transportation|
                 logistics|driver|truck|delivery") ~ 
        "Transportation and Logistics",
      TRUE ~ "Other"
    )
  )
```

```{r}
# Count the number of responses per desired career interest 
desired_careers <- intake |> 
  count(desired_careers, sort = TRUE)

# Plot the horizontal bar chart
ggplot(desired_careers, aes(x = reorder(desired_careers, n), y = n, fill = desired_careers)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(position='right') +
  coord_flip() +
  labs(
    title = "Distribution of Desired Careers by Category",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
# View all items categorized as "Other"
other_items <- intake |> 
  filter(desired_careers == "Other") |> 
  select(desired_career_interests) |> 
  distinct()
other_items
```

# TRAINING$EDUCATION DATASET

## Visualize training outcomes and career clusters
```{r}
#Training Outcomes and Career Clusters
# Remove missing values for training_status
training_edu_clean <- training_edu |> 
  filter(!is.na(training_status))

# Count training_status and calculate percentages
training_status_summary <- training_edu_clean |> 
  count(training_status, sort = TRUE) |> 
  mutate(
    percent = round(n / sum(n) * 100, 0),
    label = paste0(percent, "%")
  )

# Remove missing values for career_cluster
training_edu_clean <- training_edu |> 
  filter(!is.na(career_cluster))

# Count career_cluster and calculate percentages
career_cluster_summary <- training_edu_clean |> 
  count(career_cluster, sort = TRUE) |> 
  mutate(
    percent = round(n / sum(n) * 100, 0),
    label = paste0(percent, "%")
  )

# Pie chart
ggplot(training_status_summary, aes(x = "", y = n, fill = training_status)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) +
  labs(title = "Training Completion Status") +
  theme_void() +
  theme(legend.title = element_blank())

```

## Career_cluster

```{r}
# Remove missing values for career_cluster
training_edu_clean <- training_edu |> 
  filter(!is.na(career_cluster))

# Count career_cluster and calculate percentages
career_cluster_summary <- training_edu_clean |> 
  count(career_cluster, sort = TRUE) |> 
  mutate(
    percent = round(n / sum(n) * 100, 0),
    label = paste0(percent, "%")
  )

# Pie chart
ggplot(career_cluster_summary, aes(x = "", y = n, fill = career_cluster)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Training Completion Career Cluster") +
  theme_void() +
  theme(legend.title = element_blank())
```


# EMPLOYMENT DATASET
## Top Occupations and Hourly Wages

```{r}
#Select variables of interest
employment <- employment |> 
  select(response_id, position_job_title, job_zone, job_classification,  
         new_wage_amount_hourly, hours_per_week, career_cluster_1, 
         employment_status_all_positions)
employment
```

```{r}
# Standardize similar position_job_titles (overwrites employment clomun)
employment <- employment |>
  mutate(
    position_job_title = case_when(
      # Security 
      str_detect(tolower(position_job_title), "security") ~ "Security",
      
      # Customer Service 
      str_detect(str_to_lower(position_job_title), "customer service|customer support|call center|call centre|call center rep|call centre rep") ~
        "Customer Service",
      
      # Food prep 
      str_detect(tolower(position_job_title), "food prep|cava") ~ "Food Prep",
      
      # Housekeeping 
      str_detect(tolower(position_job_title), "housekeeping") ~ "Housekeeping",
      
      # Keep everything else as-is
      TRUE ~ position_job_title
    )
  )


employment |>
  count(position_job_title, sort = TRUE)

```


# Top 20 highest paying jobs
```{r}
# Remove rows with missing wages
employment <- employment |>
  filter(!is.na(new_wage_amount_hourly))

# Summarize wages by position_job_title
wage_summary <- employment |>
  group_by(position_job_title) |>
  summarise(
    count = n(),
    mean_wage = mean(new_wage_amount_hourly),
    median_wage = median(new_wage_amount_hourly),
    min_wage = min(new_wage_amount_hourly),
    max_wage = max(new_wage_amount_hourly)
  ) |>
  arrange(desc(mean_wage)) |>
  slice_head(n = 20) # select only top 20


# Plot top 20 positions by mean wage
ggplot(wage_summary, aes(x = reorder(position_job_title, mean_wage), y = mean_wage)) +
  geom_col(fill = "maroon") +
  scale_y_continuous(position='right', labels = dollar) +
  coord_flip() +
  labs(
    title = "Top 20 Current Titles by Hourly Wages",
    x = "",
    y = ""
  ) +
  theme_minimal()
```

## Employment Type
```{r}
#Full-Time vs Part-Time
employment |> 
  count(job_classification) |> 
  ggplot(aes(x = job_classification, y = n, fill = job_classification)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Job Classification",
       x = "Type",
       y = "") +
  theme_minimal()
```


# JULY DATASET
## Top Desired Occupations

**Overall Career Goals in July**

```{r}
# Drop columns with > 60% missing in monthly data
threshold <- 0.6 * nrow(july_data)
july_data <- july_data |> 
  select(where(~ sum(is.na(.)) < threshold))

# Select and clean monthly data
july_data <- july_data |> 
  select(response_id, educational_goal,educational_goal_status, percentage_progress_towards_career_goal,overall_career_goal, career_goal_pathway,career_goal_occupation_title, career_focus_classification) |>
  mutate(across(where(is.character), as.factor)) |>
  mutate(across(where(is.factor), ~ fct_explicit_na(., na_level = "Unknown")))
july_data
```

```{r}
#Overall Career Goals in July
career_goal_long <- july_data |>
  filter(!is.na(overall_career_goal)) |>
  mutate(
    # split line break between goals
    goal_raw = str_split(overall_career_goal, "\\r\\n")
  ) |>
  unnest(goal_raw) |>
  mutate(goal_raw = str_trim(goal_raw))

# clean label
career_goal_long <- career_goal_long |>
  mutate(
    goal_simple = case_when(
      goal_raw %in% c("1.Obtain employment")                           ~ "Obtain employment",
      goal_raw %in% c("2.Shift into a new career path")                ~ "Shift into new career",
      goal_raw %in% c("3.Advance to a leadership position")            ~ "Advance to leadership",
      goal_raw %in% c("4.Start own business")                          ~ "Start own business",
      goal_raw %in% c("5.Earn a promotion")                            ~ "Earn a promotion",
      goal_raw %in% c("6.Other Goal", "6. Other Goal")                 ~ "Other goal",
      goal_raw %in% c("7.Other Financial")                             ~ "Other financial",
      goal_raw %in% c("Unknown")                                       ~ "Unknown",
      TRUE                                                             ~ "Other/uncoded"
    )
  )

# Summarize counts and percentages
career_goal_summary <- career_goal_long |>
  count(goal_simple, sort = TRUE) |>
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label   = paste0(percent, "%")
  )

# Bar chart of overall career goals
ggplot(career_goal_summary,
       aes(x = reorder(goal_simple, percent),
           y = percent)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = label), hjust = 0.2, size = 4) +
  scale_y_continuous(position='right') +
  coord_flip() +
  labs(
    title = "Overall Career Goals (Month of July)",
    x = "",
    y = ""
  ) +
  theme_minimal()
```


# Education Goals in the month of July
```{r}
#Overall Education Goals in July
# Define keywords for educational goal
july_goal <- july_data |>
  mutate(
    educational_goal = case_when(
      str_detect(educational_goal %||% "", "1.Obtain GED/high school diploma") ~ "GED/high school diploma",
      str_detect(educational_goal %||% "", "2.Obtain career license/certification") ~ "License/Certification",
      str_detect(educational_goal %||% "", "3.Obtain bachelor's degree") ~ "Bachelor's degree",
      str_detect(educational_goal %||% "", "4.Obtain associate degree") ~ "Associate degree",
      str_detect(educational_goal %||% "", "5.Obtain graduate degree") ~ "Graduate degree",
      str_detect(educational_goal %||% "", "6.Gain general career skills") ~ "Gain general career skills",
      str_detect(educational_goal %||% "", "7.Other goal") ~ "Other goal",
      str_detect(educational_goal %||% "", "8.No educational goal") ~ "No educational goal",
      TRUE ~ "Other goal"
    )
  )

# Count and plot
july_goal <- july_goal|>
  count(educational_goal, sort = TRUE)
```

```{r}
ggplot(july_goal, aes(x = reorder(educational_goal, n), y = n)) +
  geom_col(fill = "royalblue") +
  coord_flip() +
  scale_y_continuous(position = "right") +
  labs(
    title = "Distribution of Educational Goals (month of July)",
    x = "Education Goal",
    y = ""
  ) +
  theme_minimal()
```

# Education Goal Completion Status 
```{r}
#Education Goal Completion Status 
# Clean and summarize educational goal status for plotting
goal_status_summary <- july_data |>
  filter(!is.na(educational_goal_status)) |>
  mutate(
    educational_goal_status = case_when(
      educational_goal_status %in% c("In progress", "In Progress") ~ "In Progress",
      TRUE ~ educational_goal_status
    )
  ) |>
  count(educational_goal_status, sort = TRUE) |>
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label = paste0(percent, "%")
  )
```

```{r}
# Bar charts of education goals status 
ggplot(goal_status_summary,
             aes(x = educational_goal_status, y = percent, fill = educational_goal_status)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = label), vjust = -0.3, size = 4) +
  labs(
    title = "Educational Goal Status",
    x = "",
    y = "Percentage of participants"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


# Career Focus Classification

```{r}
#Career Focus Classification
july_data |> 
  count(career_focus_classification)
# Summarize career_focus_classification and create short group labels
career_focus_summary <- july_data |>
  filter(!is.na(career_focus_classification)) |>
  count(career_focus_classification, sort = TRUE) |>
  mutate(
    percent = round(n / sum(n) * 100, 1),
    label   = paste0(percent, "%"),
    # axis labels
    career_focus_group = case_when(
      career_focus_classification == "Group 1 – Career is the Primary Focus"            ~ "Grp 1",
      career_focus_classification == "Group 2 – Career and Stability Dual Focus"        ~ "Grp 2",
      career_focus_classification == "Group 3 – Stability Focus"                        ~ "Grp 3",
      career_focus_classification == "Group 4 – Permanent Housing Referral"             ~ "Grp 4",
      career_focus_classification == "Unknown"                                          ~ "Unknown"
    ),
    career_focus_group = factor(
      career_focus_group,
      levels = c("Grp 1", "Grp 2", "Grp 3", "Grp 4", "Unknown")
    )
  )

```

```{r, echo=FALSE}
ggplot(career_focus_summary,
       aes(x = career_focus_group,
           y = percent,
           fill = career_focus_group)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = label), vjust = -0.3, size = 4) +
  scale_fill_discrete(
    name = "Career focus group",
    labels = c(
      "Grp 1"   = "Group 1 – Career is the Primary Focus",
      "Grp 2"   = "Group 2 – Career and Stability Dual Focus",
      "Grp 3"   = "Group 3 – Stability Focus",
      "Grp 4"   = "Group 4 – Permanent Housing Referral",
      "Unknown" = "Unknown"
    )
  ) +
  labs(
    title = "Career Focus Classification",
    x = "",
    y = "Percentage of participants"
  ) +
  theme_minimal()
```

# Standardize career goal occupation title 

```{r, include=FALSE}
# Career Goal Occupation Title 
patterns <- list(
  "Social Work" = c("social", "case", "human services", "custodial", "community outreach"),
  "Dental Hygienist" = c("dental hygienist", "pharmacy technician"),
  "Nursing" = c("nurs", "surgical", "phlebot", "sonograph", "respiratory", 
                "radiology", "midwife", "therapist", "mental health", "scientist", "labor/delivery physician", "veternarian"),
  "Healthcare Support" = c("assistant", "aide", "medical", "coder", "billing", "veterinary", "health information management"),
  "Education" = c("teacher", "educat", "paraprofessional", "school"),
  "Childcare" = c("child", "cda"),
  "Entrepreneurship" = c("entrepreneur", "owner", "salon", "cater", "business", "entrepeneur", "enterpreneur"),
  "Financial Services" = c("finance", "fundraising", "account", "insurance", "payroll", "financial advisor", "financial hr professional"),
  "Peer Support" = c("peer", "mental health"),
  "Maintenance" = c("plumb", "maintenance", "custodial", "technician", "environmental", 
                    "housekeep", "dishwasher", "maintenace worker", "waste management"),
  "Legal" = c("attorney", "paralegal", "judge"),
  "Real Estate" = c("real estate", "realtor", "leasing", "architectural"),
  "IT Professionals" = c("it", "cyber", "technician", "microsoft", "google", "comptia", "cisco"),
  "Customer Service" = c("call center", "customer service", "dispatcher", "receptionist", "tsa"),
  "Law Enforcement" = c("mpd", "police", "law enforce", "marshal", "security officer", "crime", "sergeant", "superintendent"),
  "Security Guards" = c("security guard", "armed security", "unarmed security", "protective officer"),
  "Administrative" = c("admin", "clerical", "office", "human resource"),
  "Chef/ Food Related" = c("chef", "cook", "server", "restaurant", "food", "baker"),
  "Program Management" = c("program specialist", "program manager", "director", "contractor"),
  "Drivers" = c("cdl", "driver", "bus operator", "truck", "motor vehicle operator"),
  "Cosmetology" = c("cosmetolog", "esthetician", "barber", "massage", "groomer"),
  "Retail Management" = c("manager", "director", "supervisor", "sales associate", "management"),
  "Federal/ Govt Work" = c("usps", "federal", "fema", "gov"),
  "Hospitality" = c("hotel", "hospitality", "funeral"),
  "Special Fields" = c("engineer", "firefighter", "vendor machine operator"),
  "Undecided" = c("undecided", "^na$", "aspiring", "instructor")
)

# Collapse patterns into regex strings
patterns <- map(patterns, ~ paste(.x, collapse = "|"))

# Apply classification
july_data <- july_data |> 
  mutate(
    title_lower = tolower(career_goal_occupation_title),
    title_lower = if_else(is.na(title_lower), "", title_lower),   # replace NA with empty string

    career_goal_occupation_title_hold = map_chr(
      title_lower,
      ~ {
        found <- NA_character_
        for (cat in names(patterns)) {
          if (isTRUE(str_detect(.x, patterns[[cat]]))) {
            found <- cat
            break
          }
        }
        ifelse(is.na(found), "Other", found)
      }
    )
  ) |> 
  select(-title_lower)
```


# Desired Career Clusters

```{r,echo=FALSE}
#Desired Career Clusters
# Count number of clients in each career goal category
career_counts <- july_data |> 
  count(career_goal_occupation_title_hold, sort = TRUE)

# Keep only top 10 categories
top_careers_clusters <- career_counts |> 
  slice_max(n, n = 10)

# Horizontal bar chart
ggplot(top_careers_clusters, aes(x = reorder(career_goal_occupation_title_hold, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  scale_y_continuous(position='right') +
  coord_flip() +  # Horizontal bars
  labs(
    title = "Top Career Goals (July 2025)",
    x = "",
    y = ""
  ) +
  theme_minimal()
```


# Current Job Title, Hourly Wage and Market Wage and wage gap
```{r, include=FALSE}
#Where they are currently working, how much they are being paid and how much the market rate is for the same job. 
wage_comparison <- tribble(
  ~current_title, ~job_zone, ~hourly_rate, ~market_hourly_rate, ~market_annual_rate, ~wage_gap, ~projected_growth,
  "Owner", 5, 62.50, 115, 239200, 52.2, "Average – 3% - 4%",
  "Case Manager", 4, 46, 32.42, 67440, -13.58, "Faster than average – 5% - 6%",
  "Security", 2, 33.56, 30.35, 63130, -3.21, "Little or no change",
  "Customer Service", 2, 46, 23.66, 49210, -22.34, "Decline -1% or lower",
  "Housekeeping", 1, 21, 18.74, 38900, -2.26, "Little or no change",
  "Book Seller", 2, 20, 18.17, 37800, -1.83, "Little or No change",
  "Cook", 2, 24, 21.75, 45230, -2.25, "Much faster than average 7% or higher",
  "Site Supervisor", 3, 20, 51.47, 107060, 31.47, "Average – 3% - 4%",
  "Cook", 2, 24, 21.75, 45230, -2.25, "Much faster than average 7% or higher",
  "Home Health Aide", 2, 19, 18.95, 39420, -0.05, "Much faster than average 7% or higher",
  "Mail Carrier", 2, 19, 28.07, 58390, -9.07, "Decline -1% or lower"
)
```

```{r, echo=FALSE}
# Wage_comparison and gap are numeric
wage_comparison <- wage_comparison |>
  mutate(
    hourly_rate = as.numeric(hourly_rate),
    market_hourly_rate = as.numeric(market_hourly_rate),
    wage_gap = as.numeric(wage_gap)
  )

# Order by market rate
wage_comparison <- wage_comparison |>
  arrange(market_hourly_rate)

# Long format for current vs market
wage_long <- wage_comparison |>
  select(current_title, hourly_rate, market_hourly_rate, wage_gap) |>
  pivot_longer(
    cols = c(hourly_rate, market_hourly_rate),
    names_to = "indicator",
    values_to = "value"
  )

# Clean labels
wage_long <- wage_long |>
  mutate(
    indicator = dplyr::recode(
      indicator,
      "hourly_rate" = "Current hourly rate",
      "market_hourly_rate" = "Market hourly rate"
    )
  )

dodge_width <- 0.7

# Plot 
ggplot(wage_long,
       aes(x = reorder(current_title, value),
           y = value,
           fill = indicator)) +
  # paired bars
  geom_col(position = position_dodge(width = dodge_width), width = 0.6) +
  # wage-gap labels positioned at current rate
  geom_text(
    data = wage_comparison,
    aes(
      x = current_title,
      y = hourly_rate,  
      label = paste0(
        ifelse(wage_gap >= 0, "+", ""),
        round(wage_gap, 1)
        )
      ),
  inherit.aes = FALSE,
  position = position_dodge(width = 0.7), 
  hjust = 1,             
  vjust = -0.5,          
  size = 3.5,
   ) +

  coord_flip() +
  scale_y_continuous(labels = dollar_format(), position = "right") +
  scale_fill_manual(values = c(
    "Current hourly rate" = "darkgreen",
    "Market hourly rate"  = "darkorange"
  )) +
  labs(
    title = "Current vs Market Hourly Rate and Wage Gap",
    subtitle = "Gap for top 10 current titles (Market - Current) hourly rate difference",
    x = "",
    y = "",
    fill = ""
  ) +
  theme_minimal() 
```

# Join Employment Data and Monthly (July) data
## Career goals among clients with the highest-paying jobs
```{r, include=FALSE}
# Select only the columns position title and hourly wage from employembt data
employment_subset <- employment |> 
  select(response_id, position_job_title, new_wage_amount_hourly) 

# Join july_data with employment using inner join to keep only matching rows between two datasets
july_employment <- july_data |> 
  inner_join(employment_subset, by = "response_id") |>
  arrange(desc(new_wage_amount_hourly))

# Check the resulting dataset
names(july_employment)
july_employment
```

```{r, include=FALSE}
# Standardize job titles and collapse similar roles into broader categories
career_counts <- july_employment |>
  select(response_id, career_goal_occupation_title,
         career_goal_occupation_title_hold,
         position_job_title,
         new_wage_amount_hourly) |>
  filter(
    !is.na(career_goal_occupation_title_hold),
    !is.na(position_job_title),
    !is.na(new_wage_amount_hourly)
  ) |>
  mutate(
    position_job_title = str_to_lower(position_job_title),
    position_job_title = case_when(
      str_detect(position_job_title, "customer|call|agent") ~ "Customer Service",
      str_detect(position_job_title, "security|armed|protective") ~ "Security",
      str_detect(position_job_title, "food|chef|cook|cava") ~ "Food Prep",
      str_detect(position_job_title, "suite|front desk|night auditor") ~ "Hospitality",
      str_detect(position_job_title, "admin|administrator") ~ "Administration",
      str_detect(position_job_title, "cashier|deli|produce|stocker") ~ 
        "Retail / Food Service",
      TRUE ~ str_to_title(position_job_title)
    )
  )

# Identify the top 10 highest-paying job categories
top_positions <- career_counts |>
  group_by(position_job_title) |>
  summarise(avg_hourly_wage = mean(new_wage_amount_hourly, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(avg_hourly_wage)) |>
  head(10)  

#Count number of clients per career goal within each job category
top_careers_10 <- career_counts |>
  filter(position_job_title %in% top_positions$position_job_title) |>
  count(position_job_title, career_goal_occupation_title_hold)


top_careers_10 <- top_careers_10 |>
  mutate(position_job_title = factor(position_job_title,
                                     levels = top_positions$position_job_title))

```

#Career Clusters
```{r, echo=FALSE}
# Plot career goals for top 10 job categories
ggplot(top_careers_10,
       aes(x = reorder(career_goal_occupation_title_hold, n),
           y = n,
           fill = position_job_title)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Career Goal Clusters for Top 10 Highest-Paying Jobs",
    subtitle = "Distribution of career goals among clients with the highest-paying jobs",
    x = "Career Goals",
    y = "",
    fill = "Current Job Title"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 10)
  )
```


# Career Goals they actually want to pivot to. 
```{r, echo=FALSE}
# Career Goals for Top 10 Highest Paying Current Job Titles

career_data <- july_employment |>
  select(response_id,
         career_goal_occupation_title,
         position_job_title,
         new_wage_amount_hourly) |>
  filter(
    !is.na(career_goal_occupation_title),
    !is.na(position_job_title),
    !is.na(new_wage_amount_hourly)
  ) |>
  mutate(
    
    position_job_title = str_squish(str_to_title(position_job_title)),
    career_goal_occupation_title = str_squish(str_to_title(career_goal_occupation_title)),
    
    # Combine similar career goals
    career_goal_occupation_title = case_when(
      str_detect(career_goal_occupation_title, regex("^Cdl$", ignore_case = TRUE)) ~ 
        "CDL Bus Driver",
      str_detect(career_goal_occupation_title, regex("Cdl Bus Driver", 
                                                     ignore_case = TRUE)) ~ "CDL Bus Driver",
      str_detect(career_goal_occupation_title, 
                 regex("Restaurant Owner|Food Truck Owner|Daycare Owner", 
                       ignore_case = TRUE)) ~ "Entrepreneur",
      str_detect(career_goal_occupation_title, regex("Emt", ignore_case = TRUE)) ~ 
        "Emergency Medical Technician",
      str_detect(career_goal_occupation_title, regex("It Specialist", ignore_case = TRUE)) ~ 
        "IT Specialist",
      TRUE ~ career_goal_occupation_title  # Keep everything else as-is
    )
  )

# Identify top 10 highest-paying current job titles
top_positions <- career_data |>
  group_by(position_job_title) |>
  summarise(avg_hourly_wage = mean(new_wage_amount_hourly, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(avg_hourly_wage)) |>
  head(10)  

# Filter data to include only top 10 current jobs
top_careers <- career_data |>
  filter(position_job_title %in% top_positions$position_job_title) |>
  count(position_job_title, career_goal_occupation_title)

# Make position_job_title a factor to preserve top 10 order
top_careers <- top_careers |>
  mutate(position_job_title = factor(position_job_title,
                                     levels = top_positions$position_job_title))

# Plot
ggplot(top_careers,
       aes(x = reorder(career_goal_occupation_title, n),
           y = n,
           fill = position_job_title)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Goals for Top 10 Highest Paying Current Job Titles",
    subtitle = "Actual occupations clients want to move to",
    x = "Desired Career Occupation",
    y = "",
    fill = "Current Job Title"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 10)
  )
```


# Career Growth opportunities, 2 per job title
```{r}

career_growth <- tibble(
  current_role = c(
    "Owner", "Owner",
    "Case Manager", "Case Manager",
    "Security Guard", "Security Guard",
    "Customer Service Representative", "Customer Service Representative",
    "Housekeeper", "Housekeeper",
    "Book Seller", "Book Seller",
    "Site Supervisor", "Site Supervisor",
    "Food Prep", "Food Prep",
    "Home Health Care Aide", "Home Health Care Aide",
    "Mail Carrier", "Mail Carrier", "Mail Carrier"
  ),
  
  current_wage = c(
    62.50, 62.50,
    46, 46,
    33.56, 33.56,
    23.57, 23.57,
    21, 21,
    20, 20,
    20, 20,
    24, 24,
    10, 10,
    19, 19, 19
  ),
  
  transition_job = c(
    "Compliance Managers", "Chief Executives",
    "Clinical and Counseling Psychologists", "Medical and Health Services Managers",
    "First-Line Supervisors of Security Workers", "Security Management Specialists",
    "Sales Representatives (Non-Advertising/Insurance/Finance/Travel)", "Sales Managers",
    "Lodging / Hospitality Managers", "First-Line Supervisors of Personal Service Workers",
    "Library Assistant", "File Clerk",
    "Solar Energy Installation Managers", "Construction Manager",
    "Chefs and Head Cooks", "Food Service Director",
    "Nursing Assistants", "Licensed Practical / Vocational Nurses",
    "Cargo and Freight Agents", "Light Truck Drivers",
    "Transportation, Storage, and Distribution Managers"
  ),
  
  education_required = c(
    "Bachelor’s Degree", "Master’s Degree",
    "Master’s Degree + Postdoctoral Training", "Bachelor’s / Master’s Degree",
    "High School + Training", "Bachelor’s / Vocational / On-the-job",
    "High School Diploma", "Bachelor’s + Training",
    "Bachelor’s / Training", "Vocational / Associate Degree",
    "High School Diploma", "High School / Associate Degree",
    "High School + On-the-job Training", "Bachelor’s Degree",
    "Vocational / Apprenticeship / Associate Degree", "High School Diploma / Apprenticeship",
    "Certification", "Certification",
    "Bachelor’s / Apprenticeship", "High School Diploma",
    "Bachelor’s Degree"
  ),
  
  median_wage = c(
    82.65, 115.00,
    46.07, 77.46,
    37.64, 51.35,
    37.50, 74.19,
    35.14, 27.16,
    25.43, 27.67,
    46.31, 61.91,
    35.58, 38.58,
    22.53, 33.86,
    23.99, 22.89,
    69.32
  ),
  
  growth_outlook = c(
    "Average", "Average",
    "Much faster than average", "Much faster than average",
    "Average", "Average",
    "Average", "Faster than average",
    "Average", "Much faster than average",
    "Decline", "Decline",
    "Faster than average", "Much faster than average",
    "Much faster than average", "Faster than average",
    "Slower than average", "Average",
    "Much faster than average", "Much faster than average",
    "Faster than average"
  )
)
career_growth
```

```{r, warning=FALSE}
# Table with kableExtra
career_growth |> 
  kable(format = "markdown", caption = "## Career Growth Opportunities", digits = 2) |> 
  kable_styling(full_width = FALSE, position = "center") |> 
  row_spec(0, bold = TRUE, background = "#D3D3D3") |> 
  column_spec(6, bold = TRUE) |>  # Highlight Growth Outlook column
  pack_rows("Owner", 1, 2) |> 
  pack_rows("Case Manager", 3, 4) |> 
  pack_rows("Security Guard", 5, 6) |> 
  pack_rows("Customer Service Representative", 7, 8) |> 
  pack_rows("Housekeeper", 9, 10) |> 
  pack_rows("Book Seller", 11, 12) |> 
  pack_rows("Site Supervisor", 13, 14) |> 
  pack_rows("Food Prep", 15, 16) |> 
  pack_rows("Home Health Care Aide", 17, 18) |> 
  pack_rows("Mail Carrier", 19, 21) |> 
  row_spec(which(career_growth$Growth_Outlook == "Much faster than average"), bold = TRUE, color = "green") |> 
  row_spec(which(career_growth$Growth_Outlook == "Faster than average"), 
           bold = TRUE, color = "darkgreen") |> 
  row_spec(which(career_growth$Growth_Outlook == "Average"), color = "orange") |> 
  row_spec(which(career_growth$Growth_Outlook == "Slower than average"), color = "red") |> 
  row_spec(which(career_growth$Growth_Outlook == "Decline"), color = "red", italic = TRUE)
```

## END 